---
title: "MESOMICS: test-sample MOFA workflow"
output:
  rmarkdown::html_vignette:
    toc: true  # Adds table of contents for easier navigation
    css: NULL  # Or point to a custom CSS file if needed
    self_contained: true
    
vignette: >
  %\VignetteIndexEntry{MESOMICS: test-sample MOFA workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = TRUE, warning = TRUE,
  eval = TRUE,
  fig.width=12, fig.height=8, fig.align='center'
)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

```{css, echo=FALSE}
body {
  overflow-x: auto;
  overflow-y: auto;
  width: 100%;
  max-width: 1200px;
  margin: 0 auto;
  padding: 20px;
  box-sizing: border-box;
}
#TOC {
  position: relative;
  float: none;
  width: 100%;
  margin-bottom: 20px;
  padding: 10px;
  border-bottom: 1px solid #ddd;
}
.main-container {
  clear: both;
}  /* 

/* SCROLLBARS: apply to all code/output blocks */
pre.output,
pre.message,
pre.warning,
pre,
pre.sourceCode,
div.sourceCode {
  max-height: 250px !important;  /* show vertical scrollbar when taller than 250px */
  overflow-y: auto !important;   /* vertical scroll */
  overflow-x: auto !important;   /* horizontal scroll if a line is too long */
}

```

# Overview

This vignette shows how to:

1. Add a test sample to MESOMICS MOFA inputs,
2. Run MOFA2 and ParetoTI,
3. Plot per-model test sample.

## Step 0: Set Up Python Environment

Before using MESOMICS, make sure you have the required Python environment.

Open your terminal and run:

```{bash, eval=FALSE}
conda env create -f mofa_env.yml
conda activate mofa_env
which python
# For example: /home/user/miniconda3/envs/mofa_env/bin/python
```

Example data paths

```{r}
mofa_dir <- system.file("extdata", package = "MESOMICS")
meso_csv <- system.file("extdata", "MESOMICS_latent_factors.csv", package = "MESOMICS")
test_csv <- system.file("extdata", "test-normalised-gene_count.csv", package = "MESOMICS")
meso_gc <- system.file("extdata", "Mesomics-normalised-gene_count.csv", package = "MESOMICS")

stopifnot(meso_csv != "", test_csv != "")


out_dir <- file.path('.', "output")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out_dir

Sys.setenv(MESOMICS_PYTHON = "/home/lipikal/miniconda3/envs/mofa_env/bin/python") #replace this with your 'which python' output
python_path <- Sys.getenv("MESOMICS_PYTHON", unset = NA)


```


## Step 1: add_sample_to_mofa()
```{r}
print('Starting Step 1')
MESOMICS::add_sample_to_mofa(
  test_matrix_path = test_csv,
  mofa_dir         = mofa_dir,
  value_data_types = "D_exprB_MOFA",
  outdir           = out_dir, 
  python_bin       = python_path
)
```
 
## Step 2: run_mofa()
```{r}
print('Starting Step 2')
MESOMICS::run_mofa(
  inputs_dir       = out_dir,   # <— where Step 1 wrote .RData
  outdir           = out_dir,  # <— where to write MOFA-<sample>.hdf5
  python_bin       = python_path
)
```

## Step 3: Per-model plots
```{r}
print('Starting Step 3')
MESOMICS::plot_test_samples(
  models_dir   = out_dir,     
  MESOMICS.LFs = meso_csv,
  python_bin   = python_path
)

```

**OR**

## Pipeline wrapper for Step 1 to 3
```{r}
MESOMICS::run_1to3(
  test_matrix_path = test_csv,
  mofa_dir         = mofa_dir,
  inputs_dir       = out_dir,
  models_dir       = out_dir,
  mesomics_csv     = meso_csv,
  python_bin       = python_path
)
```

# Other utilities/functions:
## MOFA Latent factors of test samples
```{r}
test_factors_file <- file.path(out_dir, "test_samples_factors.tsv")

MESOMICS::collect_testsamples_factors(models_dir   = out_dir, 
                                      outfile   = test_factors_file
)
```

## Batch effect - Density curves
```{r, fig.width=10, fig.height=8, out.width="100%"}
MESOMICS::plot_batch_effects(meso_matrix_path = meso_gc, 
                             test_matrix_path = test_csv, 
                             python_bin = python_path)

```

## MOFA Latent Factors Violin Plots (detect if test samples' factor values are far outside the range of MESOMICS samples)
```{r, fig.width=10, fig.height=8, out.width="100%"}

MESOMICS::plot_factor_violins(meso_factors_path = meso_csv, 
                             test_factors_path = test_factors_file,
                             python_bin   = python_path

)
```








