---
title: "MESOMICS: test-sample MOFA workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MESOMICS: test-sample MOFA workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = TRUE, warning = TRUE,
  eval = FALSE
)
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
```

# Overview

This vignette shows how to:

1. Add a test sample to MESOMICS MOFA inputs,
2. Run MOFA2 and ParetoTI,
3. Plot per-model test sample.

## Step 0: Set Up Python Environment

Before using MESOMICS, make sure you have the required Python environment.

Open your terminal and run:

```{bash}
conda env create -f mofa_env.yml
conda activate mofa_env
which python
# For example: /home/user/miniconda3/envs/mofa_env/bin/python
```

Example data paths

```{r}
mofa_dir <- system.file("extdata", package = "MESOMICS")
meso_csv <- system.file("extdata", "MESOMICS_latent_factors.csv", package = "MESOMICS")
test_csv <- system.file("extdata", "test.csv", package = "MESOMICS")

stopifnot(meso_csv != "", test_csv != "")


out_dir <- file.path(tempdir(), "mesomics-vignette")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out_dir

Sys.setenv(MESOMICS_PYTHON = "/path/to/your/env/python")
python_path <- Sys.getenv("MESOMICS_PYTHON", unset = NA)


```


## Step 1: add_sample_to_mofa()
```{r}
print('Starting Step 1')
MESOMICS::add_sample_to_mofa(
  test_matrix_path = test_csv,
  mofa_dir         = mofa_dir,
  value_data_types = "D_exprB_MOFA",
  outdir           = out_dir, 
  python_bin       = python_path
)
```
 
## Step 2: run_mofa() with basilisk
```{r}
print('Starting Step 2')
MESOMICS::run_mofa(
  inputs_dir       = out_dir,   # <— where Step 1 wrote .RData
  outdir           = out_dir,  # <— where to write MOFA-<sample>.hdf5
  python_bin       = python_path
)
```

## Step 3: Per-model plots
```{r}
print('Starting Step 3')
MESOMICS::plot_test_samples(
  models_dir   = out_dir,     # <— where MOFA-*.hdf5 live
  MESOMICS.LFs = meso_csv,
  python_bin   = python_path
)

```

## All sample plot
```{r}
MESOMICS::plot_test_all_samples(
  models_dir   = out_dir,
  mesomics_csv = meso_csv,
  out_pdf      = file.path(out_dir, "plots", "all_test_samples.pdf"),
  python_bin   = python_path
)
```

## Pipeline wrapper for Step 1 to 3:
```{r}
MESOMICS::run_1to3(
  test_matrix_path = test_csv,
  mofa_dir         = mofa_dir,
  inputs_dir       = out_dir,
  models_dir       = out_dir,
  mesomics_csv     = meso_csv,
  python_bin      = python_path
)
```


